---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

Using the publically available data on police incidents in Hartford (https://data.hartford.gov/Public-Safety/Police-Incidents-01012005-to-Current/889t-nwfu) predicting the possible incidents based on ~0.5M rows.

Note: This analysis is based on similar analysis by Max Woolf (@minimaxir)

## Setting up packages

```{r}
library(lightgbm)
library(Matrix)
library(caret)
library(viridis)
library(ggmap)
library(randomcoloR)
library(dplyr)

```
## Loading data 

Importing the data and randomizing it

```{r}
df = read.csv ("C:/Users/Niharika/Desktop/Police Incidents Prediction/police_incidents.csv")
df = df[,2:7]
set.seed(123)
df <- df %>% sample_frac()
df %>% head()


```
## Feature Engineering
# Creating time based features - Year, Month, Hour

```{r}
df <- df %>% mutate(month = substring(Date, 1, regexpr('/', Date)-1),
                    hour = substring(Time, 1, regexpr(':', Time)-1),
                    year = as.numeric(substring(Date, (nchar(as.character(Date))- 3),nchar(as.character(Date)))))
df %>% head()
```

## Changing day of the week to factor and also encoding the order

```{r}
dow_order <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
df <- df %>%
        mutate(DayOfWeek = factor(DayOfWeek, levels=dow_order))
df %>% select(DayOfWeek) %>% head()
```
Mapping category to index starting with 0

```{r}
df <- df %>%
        mutate(category_index = as.numeric(factor(Category)) - 1)
df %>% select(category_index, Category) %>% head()
```
## Training lightgbm model

Using caret to split into train/test to ensure balanced distribution of categories between test and train

```{r}
# declare categorical feature names
categoricals <- NULL
# proportion of data to train on
split <- 0.7
set.seed(123)
trainIndex <- createDataPartition(df$category_index, p= split, list = FALSE, times = 1)
dtrain <- lgb.Dataset((df %>% select(X, Y, hour, month, year, DayOfWeek) %>% data.matrix())[trainIndex,],
                     colnames = c("X", "Y", "hour", "month", "year", "DayOfWeek"),
                     categorical_feature = categoricals,
                     label = df$category_index[trainIndex], free_raw_data=T)
dtest <- lgb.Dataset.create.valid(dtrain,
                                  (df %>% select(X, Y, hour, month, year, DayOfWeek) %>% data.matrix())[-trainIndex,],
                                  label = df$category_index[-trainIndex])
params <- list(objective = "multiclass", metric = "multi_logloss")
valids <- list(test=dtest)
num_classes <- length(unique(df$category_index))

# preformatting sizes for use in data visualizations later
train_size_format <- length(trainIndex) %>% format(big.mark=",")
test_size_format <- (df %>% nrow() - length(trainIndex)) %>% format(big.mark=",")

```



```{r}

#Training the model on train dataset
bst <- lgb.train(params,
                dtrain,
                nrounds = 500,
                valids,
                num_threads = 4,
                num_class = num_classes,
                verbose = 0,
                record = T,
                early_stopping_rounds = 5,
                categorical_feature = categoricals
                )

# multilogloss of final iteration on train
paste("# Rounds:", bst$current_iter())
paste("Multilogloss of best model:", bst$record_evals$test$multi_logloss$eval %>% unlist() %>% tail(1))

#calculating variable importance
df_imp <- tbl_df(lgb.importance(bst, percentage = TRUE))
df_imp


#running the model on test
test <- (df %>% select(X, Y, hour, month, year, DayOfWeek) %>% data.matrix())[-trainIndex,]
preds_matrix <- predict(bst, test, reshape=T)
preds_cor <- cor(preds_matrix)
results <- t(apply(preds_matrix, 1, function (x) {
  max_index = which(x==max(x))
  return (c(max_index-1, x[max_index]))
}))


df_results <- data.frame(results, label_act = df$category_index[-trainIndex]) %>%
                tbl_df() %>%
                transmute(label_pred = X1, prod_pred = X2, label_act)

rm(preds_matrix)


```


Building the confusion matrix


```{r}
cm <- confusionMatrix(df_results$label_pred, df_results$label_act)
data.frame(cm$overall)

```
##Visualizations

Importance Bars

```{r}
df_imp$Feature <- factor(df_imp$Feature, levels = rev(df_imp$Feature))

plot <- ggplot(df_imp, aes(x=Feature, y=Gain)) +
          geom_bar(stat="identity", fill="#34495e", alpha=0.9) +
          geom_text(aes(label=sprintf("%0.1f%%", Gain*100)), color="#34495e", hjust=-0.25,family = "Times" ,size=2.5) +
          fte_theme() +
          coord_flip() +
          scale_y_continuous(limits = c(0, 0.4), labels=percent) +
   theme(plot.title=element_text(hjust=0.5), axis.title.y=element_blank()) +
          labs(title="Feature Importance for Hartford Arrest Type Model", y="% of Total Gain in LightGBM Model")

plot

```
Cofusion Matrix

```{r}
df_cm <- tbl_df(data.frame(cm$table))
#creating mapping
df_labels <- df %>%
              select(category_index, Category) %>%
              group_by(category_index, Category) %>%
              summarize() %>%
              ungroup() %>%
              mutate(category_index = factor(category_index))
df_cm <- df_cm %>%
                left_join(df_labels, by = c("Prediction" = "category_index")) %>%
                left_join(df_labels, by = c("Reference" = "category_index")) %>%
                rename(label_pred = Category.x, label_act = Category.y)

df_correct <- df_cm %>% filter(label_pred == label_act)
plot <- ggplot(df_cm, aes(x=label_act, y=label_pred, fill = Freq)) +
          geom_tile() +
          geom_point(data=df_correct, color="white", size=0.8) +
          fte_theme() +
          coord_equal() +
          scale_x_discrete() +
          scale_y_discrete() +
          theme(legend.title = element_text(size=7, family="Open Sans Condensed Bold"), legend.position="top", legend.direction="horizontal", legend.key.width=unit(1.25, "cm"), legend.key.height=unit(0.25, "cm"), legend.margin=unit(0,"cm"), axis.text.x=element_text(angle=-90, size=6, vjust=0.5, hjust=0), axis.text.y=element_text(size=6), plot.title = element_text(hjust=1)) +
            scale_fill_viridis(name="# of Preds", labels=comma, breaks=10^(0:4), trans="log10") +
            labs(title = sprintf("Confusion Matrix between %s Predicted HPD Arrest Labels and Actual", test_size_format),
                 x = "Actual Label of Arrest",
                 y = "Predicted Label of Arrest")

plot
```
Correlations



```{r}
dd <- as.dist((1-preds_cor)/2)
hc <- hclust(dd, "centroid")
label_order <- hc$order
preds_cor_reorder <- preds_cor[label_order, label_order]
df_corr <- tbl_df(data.frame(Var1=c(row(preds_cor_reorder))-1, Var2=c(col(preds_cor_reorder))-1, value = c(preds_cor_reorder))) %>%
            filter(Var1 <= Var2) %>%
            mutate(Var1 = factor(Var1), Var2=factor(Var2))
df_corr <- df_corr %>%
                left_join(df_labels, by = c("Var1" = "category_index")) %>%
                left_join(df_labels, by = c("Var2" = "category_index")) %>%
                mutate(label1 = factor(Category.x), label2 = factor(Category.y))
# fix the label order to the reordered order from the hclust
levels(df_corr$label1) <- levels(df_corr$label1)[label_order]
levels(df_corr$label2) <- levels(df_corr$label2)[label_order]
plot <- ggplot(df_corr, aes(x=label1, y=label2, fill=value)) +
          geom_tile() +
          fte_theme() +
          scale_x_discrete() +
          scale_y_discrete() +
          coord_fixed() +
          theme(legend.title = element_text(size=7, family="Open Sans Condensed Bold"), legend.position="top", legend.direction="horizontal", legend.key.width=unit(1.25, "cm"), legend.key.height=unit(0.25, "cm"), legend.margin=unit(0,"cm"), panel.margin=element_blank(), axis.text.x=element_text(angle=-90, vjust=0.5, hjust=0), axis.title.y=element_blank(), axis.title.x=element_blank(), plot.title=element_text(hjust=1, size=6)) +
            scale_fill_gradient2(high = "#2ecc71", low = "#e74c3c", mid = "white", 
   midpoint = 0, limit = c(-0.5,0.5), 
   name="Pearson\nCorrelation", breaks=pretty_breaks(8))  +
            labs(title = sprintf("Correlations between Predicted Multiclass Probabilities of %s HPD Arrest Category Labels", test_size_format))

plot

```



```{r}
bbox = c(-72.71864705,41.72214158,-72.64828835,41.80743325)
map <- get_map(location = bbox, source = "stamen", maptype = "toner-lite")

grid_size = 50
df_points <- data.frame(expand.grid(X=seq(bbox[1], bbox[3], length.out=grid_size),
                                    Y=seq(bbox[2], bbox[4], length.out=grid_size)))

df_points %>% head()

```

Predicting arrest types at each point on December 17 2017 5PM

```{r}
date_target <- as.POSIXct("2018-10-28 15:00:00")
df_points <- df_points %>%
              mutate(hour = format(date_target, "%H"),
                    month = format(date_target, "%m"),
                    year = format(date_target, "%Y"),
                    DayOfWeek = which(levels(df$DayOfWeek) == format(date_target, "%A"))) %>%
            data.matrix()

preds_matrix <- matrix(predict(bst, df_points), byrow=T, nrow(df_points), num_classes)
results <- t(apply(preds_matrix, 1, function (x) {
  max_index = which(x==max(x))
  return (c(max_index-1, x[max_index]))
}))
rm(preds_matrix)

df_results <- data.frame(X=df_points[,1], Y=df_points[,2], label=factor(results[,1]), prob=results[,2]) %>%
                tbl_df() %>%
                left_join(df_labels, by=c("label" = "category_index")) %>%
                mutate(Category = factor(Category))


plot <- ggmap(map) +
            geom_raster(data = df_results %>% filter(Category != "Other Offenses"), aes(x=X, y=Y, fill=Category), alpha=0.8, size=0) +
            coord_cartesian() +
            fte_theme() +
            scale_fill_brewer(palette = "Dark2") +
            theme(axis.text.x = element_blank(), axis.text.y = element_blank(), axis.title.x = element_blank(), axis.title.y = element_blank()) +
            theme(legend.title = element_text(size=7, family="Open Sans Condensed Bold"), legend.position="right", legend.key.width=unit(0.5, "cm"), legend.key.height=unit(2, "cm"), legend.margin=margin(1,0,1,0), plot.title=element_text(hjust=0, size=11)) +
            labs(title = sprintf("Locations of Predicted Types of Arrests in Hartford on %s",
                 format(date_target, '%B %d, %Y at%l %p')))

plot
```